The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, ``Parallel Algorithms'' lecture, 2012 winter term at TU Wien.}

### 2014-03-30 article/article.tex
add theorem for lower bound of agl and highlight that asymmetric tie-breaking and partitioning is crucial

Consider the slightly more involved \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ (to simplify the presentation, we will assume that $d$ divides $n$ evenly in the remainder of this survey). We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins $i$ for $1 \leq i \leq n$ and put the ball into the bin with smallest index $i$ among those with the same minimal load. Somewhat surprisingly, V\"ocking~\cite{VOC03} showed that such a strategy employing asymmetric tie-breaking in combination with non-uniform bin selection reduces the bound for the maximum load to
\begin{align*}
\load = \frac{m}{n} + \frac{\ln\ln n}{d  \ln \Phi_d} + O(1),
\end{align*}
w.h.p., for all $m$. (The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.) V\"ocking was also able to show that the Always-go-left strategy is optimal up to additive constants~\cite{VOC03}.

@@

\begin{theorem}
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load is at most 
\begin{align*}
\load \leq \frac{\ln\ln n}{d  \ln \Phi_d} + O(1) \quad ,
\end{align*}
w.h.p.,~for all $t$~\cite{VOC03}.
\end{theorem}

@@

\begin{compactitem}
\item $T_d(1)$ is a single node and  $T_d(2) = T_d(1)$.
\item For $3\leq k \leq d$, $T_d(k)$ is a tree with $k-1$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(1)$.
\item For $k>d$, $T_d(k)$ is a tree with $d$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(k-d)$.
\end{compactitem}
A $d$-ary AWT of order $L$ has the structure of a Fibonacci tree $T_d(d  L+1)$ and thus has $l=F_d(d L + 1) \geq \Phi_d^{d L-1}$ leaves (Eq.~\ref{eqn:capocelli}).
Furthermore, for each node we define a label $(h, i)$, where $h, i \in \mathbb{N}$, $0 \leq h \leq L$ and  $1\leq i \leq d$:
\begin{compactitem}
\item The root nodes of the Fibonacci trees $T_d(1)$ and $T_d(2)$ are assigned the labels $(0,1)$ and $(0,2)$, respectively. Thus, nodes labelled $(0,1)$ or $(0,2)$ do not have any children.
\item A node with label $(0, i)$, where $i>2$, has $i-1$ children with the labels $(0,i-1),\dots,(0,1)$.
\item A node with label $(h,i)$, where $h>0$, has $d$ children labelled $(h,i-1),\dots,(h,1)$,$(h-1,d),\dots,(h-1, i)$.
\item The root of a Fibonacci tree $T_d(d h+i)$ is labelled $(h, i)$. 
\end{compactitem}
Thus, the root node of an AWT of order $L$ is assigned the label $(L, 1)$. Note that the labels are unique for the set of children of any node $v$, but not necessarily for all nodes of the tree. 
\begin{lemma}
When using the Always-go-left algorithm to put balls into bins, a bin containing more than $L+3$ balls activates an AWT of order $L$.
\end{lemma}
\begin{proof}
The following invariant will be maintained while assigning balls to the nodes of the AWT:
\begin{invariant}
A ball $b_v$ referenced by a node $v$ with label $(h,i)$ was placed in a bin $x$ belonging to group $i$ that, at the time of insertion of the ball $b_v$, contained at least $h+3$ other balls. The ball represented by the root node may be put in a bin belonging to any group.
\end{invariant}
\noindent
Let $x$ be the bin that contains $L+4$ balls and assign the topmost ball $b_t$ in $x$ to the root node of the AWT. Thus, at the time of insertion of $b_t$, each of the $d$ locations of this ball points to a bin containing at least $L+3$ balls. These are the $d$ bins that were sampled by the Always-go-left scheme, i.e., bin $i$ belongs to the $i$'th group of bins, where each group consists of $n/d$ bins. The child node of the root with label $(L-1, i)$ is assigned the topmost ball of the $i$'th location. The assignment of balls to nodes then proceeds recursively: Given a node $v$ labelled $(h, i)$ that references a ball $b_v$ (which was put into some bin of the $i$-th group that, at the time of insertion of ball $b_v$, contained at least $h+3$ other balls), we can conclude the following:
\begin{compactitem}
\item For $1\leq j < i$, the $j$'th location of the ball $b_v$ contains at least $h+4$ balls, at the insertion time of the ball $b_v$. If the $j$-th location contained less balls than the $i$-th location, the Always-go-left scheme would have put the ball $b_v$ into the bin pointed to by location $j$. The topmost ball in location $j$ is assigned to the  child $u$ of $v$ with label $(h,j)$. 
\item For $i < j \leq d$, the $j$-th location contains at least $h+3$ balls. If location $j$ contained less, the Always-go-left scheme would have put the ball $b_v$ into the bin location $j$ points to. The topmost ball in the bin that location $j$ points to is assigned to the child node $u$ of $v$ labelled with $(h-1, j)$.

@@

\section{A bit of history}
\label{sec:historyOfGreedy}
The history of proving the remarkable property of the two-choice paradigm (see Theorem~\ref{theorem:greedy}) is in itself quite interesting. It was first proven by Azar et al.~\cite{ABKU99} for the case $m = n$. In the same paper, they claimed that the bound holds for arbitrary $m$. Unfortunately, their analysis breaks down for the heavily loaded case, where $m \in \Omega\left(n  \log n\right)$. Berenbrink at. all~\cite{BCSV06} were the first to prove that the same bound indeed holds for the heavily loaded case. Their proof is fairly complex and a simpler analysis may be found in~\cite{TW13}, which also provides a tight lower bound for the heavily loaded case. Curiously, the case $d = 2$ was implicitly proven by Karp et al. in the context of simulating PRAMs on distributed memory machines\cite{KLM92}. As discussed in Section~\ref{sec:AlwaysGoLeft}, V\"ocking~\cite{VOC03} was able to improve the bound as well as to show that the Always-go-left strategy that he introduces is optimal up to additive constants.

>>>>

Consider the slightly more involved \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ (to simplify the presentation, we will assume that $d$ divides $n$ evenly in the remainder of this survey). We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins $i$ for $1 \leq i \leq n$ and put the ball into the bin with smallest index $i$ among those with the same minimal load. Somewhat surprisingly, V\"ocking~\cite{VOC03} showed that such a strategy employing asymmetric tie-breaking in combination with non-uniform bin selection reduces the bound for the maximum load to
\begin{align*}
\load = \frac{m}{n} + \frac{\ln\ln n}{d  \ln \Phi_d} + O(1),
\end{align*}
w.h.p., for all $m$. (The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.) Berenbrink et al.~\cite{BCSV06} and V\"ocking~\cite{VOC03} were able to show that the Always-go-left strategy is optimal up to additive constants.

@@

\begin{theorem}
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load is at most 
\begin{align*}
\load \leq \frac{\ln\ln n}{d  \ln \Phi_d} + O(1) \quad ,
\end{align*}
w.h.p.,~for all $t$~\cite{VOC03}.
\end{theorem}
\noindent
V\"ocking~\cite{VOC03} also proved a matching lower bound for the lightly loaded case:
\begin{theorem}
\label{theorem:aglmLower}
For $m=n$, using an arbitrary allocation scheme that chooses $d$ bins for each ball at random (i.e., any allocation scheme in class 3) the maximum load is
\begin{align*}
\load = \frac{\ln \ln n}{d \ln \Phi_d} - O(1) \quad 
\end{align*}
\end{theorem}
\noindent
w.h.p.,~for all $t$~\cite{VOC03}. We not that the combination of asymmetric tie-breaking and and non-uniform (but independent) sampling of the $d$ bins is crucial to achieve bounds given in Theorems~\ref{theorem:aglm} and~\ref{theorem:aglmLower}.

@@

\begin{compactitem}
\item $T_d(1)$ is a single node and  $T_d(2) = T_d(1)$.
\item For $3\leq k \leq d$, $T_d(k)$ is a tree with $k-1$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(1)$.
\item For $k>d$, $T_d(k)$ is a tree with $d$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(k-d)$.
\end{compactitem}
A $d$-ary AWT of order $L$ has the structure of a Fibonacci tree $T_d(d  L+1)$ and thus has $l=F_d(d L + 1) \geq \Phi_d^{d L-1}$ leaves (Eq.~\ref{eqn:capocelli}).
Furthermore, for each node we define a label $(h, i)$, where $h, i \in \mathbb{N}$, $0 \leq h \leq L$ and  $1\leq i \leq d$:
\begin{compactitem}
\item The root nodes of the Fibonacci trees $T_d(1)$ and $T_d(2)$ are assigned the labels $(0,1)$ and $(0,2)$, respectively. Thus, nodes labelled $(0,1)$ or $(0,2)$ do not have any children.
\item A node with label $(0, i)$, where $i>2$, has $i-1$ children with the labels $(0,i-1),\dots,(0,1)$.
\item A node with label $(h,i)$, where $h>0$, has $d$ children labelled $(h,i-1),\dots,(h,1)$,$(h-1,d),\dots,(h-1, i)$.
\item The root of a Fibonacci tree $T_d(d h+i)$ is labelled $(h, i)$. 
\end{compactitem}
Thus, the root node of an AWT of order $L$ is assigned the label $(L, 1)$. Note that the labels are unique for the set of children of any node $v$, but not necessarily for all nodes of the tree. 
\begin{lemma}
When using the Always-go-left algorithm to put balls into bins, a bin containing more than $L+3$ balls activates an AWT of order $L$.
\end{lemma}
\begin{proof}
The following invariant will be maintained while assigning balls to the nodes of the AWT:
\begin{invariant}
A ball $b_v$ referenced by a node $v$ with label $(h,i)$ was placed in a bin $x$ belonging to group $i$ that, at the time of insertion of the ball $b_v$, contained at least $h+3$ other balls. The ball represented by the root node may be put in a bin belonging to any group.
\end{invariant}
\noindent
Let $x$ be the bin that contains $L+4$ balls and assign the topmost ball $b_t$ in $x$ to the root node of the AWT. Thus, at the time of insertion of $b_t$, each of the $d$ locations of this ball points to a bin containing at least $L+3$ balls. These are the $d$ bins that were sampled by the Always-go-left scheme, i.e., bin $i$ belongs to the $i$'th group of bins, where each group consists of $n/d$ bins. The child node of the root with label $(L-1, i)$ is assigned the topmost ball of the $i$'th location. The assignment of balls to nodes then proceeds recursively: Given a node $v$ labelled $(h, i)$ that references a ball $b_v$ (which was put into some bin of the $i$-th group that, at the time of insertion of ball $b_v$, contained at least $h+3$ other balls), we can conclude the following:
\begin{compactitem}
\item For $1\leq j < i$, the $j$-th location of the ball $b_v$ contains at least $h+4$ balls, at the insertion time of the ball $b_v$. If it contained less balls than the $i$-th location, the Always-go-left scheme would have put the ball $b_v$ into the bin pointed to by location $j$. The topmost ball in location $j$ is assigned to the child $u$ of $v$ with label $(h,j)$. 
\item For $i < j \leq d$, the $j$-th location contains at least $h+3$ balls. If it contained less, the Always-go-left scheme would have put the ball $b_v$ into the bin location $j$ points to. The topmost ball in the bin that location $j$ points to is assigned to the child node $u$ of $v$ labelled with $(h-1, j)$.

@@

\section{A bit of history}
\label{sec:historyOfGreedy}
The history of proving the remarkable property of the two-choice paradigm (see Theorem~\ref{theorem:greedy}) is in itself quite interesting. It was first proven by Azar et al.~\cite{ABKU99} for the case $m = n$. In the same paper, they claimed that the bound holds for arbitrary $m$. Unfortunately, their analysis breaks down for the heavily loaded case, where $m \in \Omega\left(n  \log n\right)$. Berenbrink at. all~\cite{BCSV06} were the first to prove that the same bound indeed holds for the heavily loaded case. Their proof is fairly complex but provides a tight analysis of the heavily loaded case; a simpler analysis may be found in~\cite{TW13}. Curiously, the case $d = 2$ was implicitly proven by Karp et al. in the context of simulating PRAMs on distributed memory machines\cite{KLM92}. As discussed in Section~\ref{sec:AlwaysGoLeft}, V\"ocking~\cite{VOC03} was able to improve the bound as well as to show that the Always-go-left strategy that he introduces is optimal up to additive constants.
